Approximation as Differentiation

### 2025-07-08 main/models.tex
[ tex ] failure of cartesian closure on biproduct categories

There is another construction of internal homs on $\Fam(\cat{C})$
arising from the use of fibrations for categorical logical relations
proofs, due to \citet{hermida99} (Corollary 4.12). If we assume that
$\cat{C}$ is itself cartesian closed and has all products, then we
could construct an internal hom as:
\begin{displaymath}
  (X, \partial X) \to (Y, \partial Y) = (X \to Y, \lambda f. \Pi_{x : X}.\,\partial X(x) \to \partial Y(f\,x))
\end{displaymath}
However, for the purposes of modelling differentiable programs, this
is fatally flawed in that neither of $\LatGal$ nor $\FinVect$ are
cartesian closed, and there is no way of making them so without losing
the property of having biproducts. We will use Hermida's construction
in our definability proof in \secref{definability}, where we use a
logical relations argument to show that every morphism definable in
the higher order language is also first-order definable.

@@

  \begin{salign*}
    (\biinj_X \comp \biproj_X) + (\biinj_Y \comp \biproj_Y) &= \id_{X \biprod Y}
  \end{salign*}
\end{definition}
As the name suggests, biproducts in a category are both products and
coproducts:

\begin{proposition}
  \begin{enumerate}
  \item A $\CMon$-category that has biproducts $X \oplus Y$ for all
    $X$ and $Y$ also has products and coproducts with
    $X \times Y = X + Y = X \oplus Y$.
  \item A $\CMon$-category with (co)products also has biproducts.
  \end{enumerate}
\end{proposition}

@@

>>>>

\begin{remark}
  \label{rem:hermida-exponentials}
  There is another construction of internal homs on $\Fam(\cat{C})$
  arising from the use of fibrations for categorical logical relations
  proofs, due to \citet{hermida99} (Corollary 4.12). If we assume that
  $\cat{C}$ is itself cartesian closed and has all products, then we
  could construct an internal hom as:
  \begin{displaymath}
    (X, \partial X) \to (Y, \partial Y) = (X \to Y, \lambda f. \Pi_{x : X}.\,\partial X(x) \to \partial Y(f\,x))
  \end{displaymath}
  However, for the purposes of modelling differentiable programs, this
  is fatally flawed in that neither of $\LatGal$ nor $\FinVect$ are
  cartesian closed, and there is no way of making them so without
  losing the property of having biproducts. We will use Hermida's
  construction in our definability proof in \secref{definability},
  where we use a logical relations argument to show that every
  morphism definable in the higher order language is also first-order
  definable.
\end{remark}

@@

  \begin{salign*}
    (\biinj_X \comp \biproj_X) + (\biinj_Y \comp \biproj_Y) &= \id_{X \biprod Y}
  \end{salign*}
  A zero object is an object that is both initial and terminal.
\end{definition}
As the name suggests, biproducts in a category are both products and
coproducts:
\begin{proposition}
  \begin{enumerate}
  \item A $\CMon$-category that has biproducts $X \oplus Y$ for all
    $X$ and $Y$ also has products and coproducts with
    $X \times Y = X + Y = X \oplus Y$.
  \item A $\CMon$-category with (co)products also has biproducts, and
    any initial or terminal object is a zero object.
  \end{enumerate}
\end{proposition}

@@

The following proposition shows why the alternative construction of
exponentials in $\Fam(C)$ described in \remref{hermida-exponentials}
is not suitable:
\begin{proposition}
  A $\CMon$-category that is cartesian closed has trivial homsets:
  $\cat{C}(X,Y) \cong 1$.
\end{proposition}

\begin{proof}
  $\cat{C}(X, Y) \cong \cat{C}(0 + X,Y) \cong \cat{C}(0 \times X,Y)
  \cong \cat{C}(0,X \to Y) \cong 1$.
\end{proof}
